---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
devtools::load_all()
```

[![](https://www.r-pkg.org/badges/version/varTestnlme)](https://CRAN.R-project.org/package=varTestnlme)

# varTestnlme

[varTestnlme](https://baeyc.github.io/varTestnlme/index.html) implements the likelihood ratio test (LRT) for testing the presence of random effects in linear, generalized linear and nonlinear mixed-effects model. The test can be used to answer questions of the type:

  * should a certain subset of random effects be in fact considered as fixed effects?
  * is there any random effects in the model?
  * are there any correlation between two subsets of random effects?
  
It is possible to compare two models with different random effects, provided that the random structures of the two models are **nested**.

The package works on models that were fitted using [nlme](https://CRAN.R-project.org/package=nlme),
[lme4](https://CRAN.R-project.org/package=lme4), or
[saemix](https://CRAN.R-project.org/package=saemix) packages.

## Reference
Baey C, Cournède P-H, Kuhn E, 2019. Asymptotic distribution of likelihood ratio test statistics for variance components in nonlinear mixed effects models. \emph{Computational Statistics and Data Analysis} 135:107--122 (2019), [https://doi.org/10.1016/j.csda.2019.01.014](https://doi.org/10.1016/j.csda.2019.01.014)

## Installation

Install from CRAN:
```{r, eval=FALSE}
install.packages("varTestnlme")
```

Install the development version from Github:

```{r, eval=FALSE}
# install.packages("devtools")
#devtools::install_github("baeyc/varTestnlme")

devtools::load_all()
```


## Example

### With default options

The first example is based on the `nlme` package. 

```{r}
# First, fit models from your preferred mixed-effects models package
library(nlme)
fm1.lis <- nlsList(rate ~ SSasympOff(pressure, Asym, lrc, c0), data=Dialyzer)
nm1e <- nlme(fm1.lis, random = pdSymm(Asym + lrc ~ 1))
nm0e <- nlme(fm1.lis, random = pdDiag(lrc ~ 1))

# Then, simply use the varTest function to compare the two models
# It will automatically detect the covariance structures
vte <- varTest(nm1e, nm0e, fim = "compute", pval.comp = "both", control=list(B=10))
```

It works similarly with lme4 package or saemix.

### Estimation of Fisher Information Matrix

The imiting distribution of the test statistic is a chi-bar-square distribution whose weights depend on the Fisher Information Matrix (FIM) of the model. Depending on the package used to fit the model and on your preferences, it is possible to specify different options with the argument `fim`:
 * `fim="extract"`: (the default option) to extract the FIM computed by the package used to fit the models. Please note that it might not be available in some cases. For example, for nonlinear models fitted with `lme4`. 
 * `fim="compute"`: to estimate the FIM using parametric bootstrap. Note that it can take some time to run. In this case, one must provide the bootstrap sample size with the `control` argument
 * `fim=I`: with `I` a symmetric positive definite matrix, to provide your own FIM

```{r}
# Bootstrap estimation of the FIM
varTest(nm1e, nm0e, fim="compute", pval.comp="both", control=list(B=100))
```


Another example using the `lme4` package and the Orange trees data.

```{r}
library(lme4)
dial.start <- unlist(nm1e$call$start)
names(dial.start) <- c("Asym","lrc","c0")
dial.start  <- c(Asym = 50, lrc = 0.5, c0 = 0.2)
nm1  <- nlmer(rate ~ SSasympOff(pressure, Asym, lrc, c0) ~ (0 + Asym + lrc || Subject), Dialyzer, start = dial.start)
nm0  <- nlmer(rate ~ SSasympOff(pressure, Asym, lrc, c0) ~ (0 + Asym | Subject), Dialyzer, start = dial.start)
vt <- varTest(nm1,nm0,fim="compute",pval.comp = "both",control=list(M=10,B=20))


library(nlme)
nm1e <- nlme(rate ~ SSasympOff(pressure, Asym, lrc, c0),
             random = pdDiag(Asym + lrc + c0 ~ 1),
             fixed = Asym + lrc + c0 ~ 1,
             data = Dialyzer,
             start = dial.start)
nm0e <- nlme(rate ~ SSasympOff(pressure, Asym, lrc, c0),
             random = pdDiag(Asym + c0 ~ 1),
             fixed = Asym + lrc + c0 ~ 1,
             data = Dialyzer,
             start = dial.start)
varTest(nm1e,nm0e)
```