---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
devtools::load_all()
```
# varTestnlme

Implements the likelihood ratio test (LRT) for testing the presence of random effects in linear, generalized linear and nonlinear mixed-effects model. The test can be used to answer questions of the type:

  * should a certain subset of random effects be in fact considered as fixed effects?
  * is there any random effects in the model?
  * are there any correlation between two subsets of random effects?
  
It is possible to compare two models with different random effects, provided that the random structures of the two models are **nested**.

The package works on models that were fitted using `nlme`, `lme4` or `saemix` packages.

## Installation

Install the development version from Github:

```{r}
# install.packages("devtools")
devtools::install_github("baeyc/varTestnlme")
```


## Example

```{r}
# First, fit models from your preferred mixed-effects models package
require(nlme)
fm1Indom.lis <- nlsList(conc ~ SSbiexp(time, A1, lrc1, A2, lrc2),data = Indometh)
m1 <- nlme(fm1Indom.lis,random = pdDiag(A1 + lrc1 + A2 + lrc2 ~ 1))
m0 <- nlme(fm1Indom.lis,random = pdDiag(A1 + lrc1 ~ 1))

# Then, simply use the varTest function to compare the two models
# It will automatically detect the covariance structures
varTest(m1,m0)
```


